Plume-SPH 1.0: A 3D dust-gas volcano plume model based on smoothed particle hydrodynamics

### 2016-10-01 Manuscripts_GMD.tex
add a pic

Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh-based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh-based methods in modeling of multiple phase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. Erupted material is represented by a single phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed in this model. Affect of wind is also ignored. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions. The solver of this model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by a comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plum models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.

@@

& = 0.5\sum_b [m_b \widehat{\textbf{v}_{\alpha b}} (\dfrac{p_b}{\rho_b^2} + \dfrac{p_{\alpha}}{\rho_{\alpha}^2} + \Pi_{\alpha b} - \Phi_{\alpha b}) \bigtriangledown_{\alpha}w_{\alpha b}(h_{\alpha}] 

@@

@@

n_s = (1 - n_a) + (1- n_{g0})

@@

@@

The Plume-SPH code with a user manual providing instructions for installation, running and visualization is archive in github (https://github.com/Plume-SPH/plume-sph.git). The input data for all simulations presented in this work can also be found in the repository.

@@

Data for running the test cases, including meteorological data, properties of erupted material and air, and eruption condition data can be found in source code repository. Raw data of simulation results is archived in google drive. Access permission will be given up request. Post-processed data for generating figures is provided as supplement along with the manuscripts. Please note that some of these pictures are generating by visualizing the raw data with visualization tools, there is no separate data for these figures.

>>>>

Volcanic ash, if persists in the atmosphere, will pose great threat on aircrafts and communities. VATDs (Volcanic ash transport and dispersion models) usually take output of plume models as source terms. The accuracy of source terms is crucial for VATDs. All of existing 3D (three dimensional) plume models are using mesh-based methods. SPH (Smoothed particle hydrodynamics), as a mesh free method has several advantages over mesh-based methods in modeling of multiphase free boundary flow. As an initial effort on exploiting the feasibility and advantage of SPH in volcano plume modeling we adopt a relative simpler model, a 3D dusty-gas dynamic model, targeting at capturing the more basic features of volcano plume. In this model, erupted material is assumed to be well mixed and represented by one phase while air is another phase. Immediate dynamic and thermal dynamic equilibrium between air and erupted material are assumed. Affect of wind is not considered. Several newly developed techniques in SPH are adopted and trimmed to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH method. The $SPH-\varepsilon$ turbulence model is adopted to capture sub-particle mixing. Heat exchange due to turbulence is then calculated with Reynolds analogy. Corrected formulism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementation of SPH. Core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and cross section with experimental results of JPUE (jet or plume which is ejected from a nozzle into a uniform environment). The top height of Pinatubo eruption (15 June 1991) simulated by our model is consistent with both observation and existing 3D plume models. Profiles of several integrated variables are compared with these by existing 3D plum models and further verifies the correctness of our model. Analysis on the plume evolution process illustrated that this model is able to reproduce the basic physics of plume development. The comparison also implies that turbulence model plays a significant role in 3D volcano plume modeling.

@@

& = 0.5\sum_b [m_b \widehat{\textbf{v}_{\alpha b}} (\dfrac{p_b}{\rho_b^2} + \dfrac{p_{\alpha}}{\rho_{\alpha}^2} + \Pi_{\alpha b} - \Phi_{\alpha b}) \bigtriangledown_{\alpha}w_{\alpha b}(h_{\alpha})] 

@@

\includegraphics[width=18cm]{t120_bc_proc.png}
\caption{A cross section view of the simulation domain. Subfigure a shows the mass fraction of a plume simulation. Subfigure b shows all boundary conditions: the dark blue is for velocity inlet boundary, the light blue is for pressure outlet boundary, the gray is for no-slip wall boundary condition. Ghost particle ID of All real particles are set to 100, so is 100 in b. Subfigure c shows the cross section view of domain decomposition based on SFC. The simulation is conducted on 144 processors (12 nodes with 12 cores each node), so there are 144 subdomains in total.}
\label{fig:bc_and_domain_decomp}

@@

n_s = (1 - n_a) (1- n_{g0})

@@

\label{eq:solid-frac-express2}

@@

The Plume-SPH code with a user manual providing instructions for installation, running and visualization are archived in github (https://github.com/Plume-SPH/plume-sph.git). The input data for all simulations presented in this work can also be found in the repository.

@@

Data for running the test cases, including meteorological data, properties of erupted material and air, and eruption condition data can be found in source code repository. Raw data of simulation results are archived in google drive. Access permission will be given upon request. Post-processed data for plotting figures is provided as supplement along with the manuscript. Please note that some of these pictures are generated by visualizing the raw data with visualization tools, there is no separate data for these figures.
